Markdown for Reproducible Research - Assignment 1
=================================================
 
This markdown is for the assignment 1 for the Course Reproducible Research as part of Data Science

Loading and preprocessing the data
----------------------------------

1.  Check if the file exists, else download and unzip and read the data

```{r}
dlZip <- "repdata_data_activity.zip"
dlTxt <- "activity.csv"

if(!file.exists(dlTxt))
{
  message("CSV file does not exist.. try for zip file")
  if(!file.exists(dlZip)){
    message("zip file does not exist as well... downloading...")
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",dlZip )
  }
  message("extracting text file from zip")
  unzip(dlZip)
}  
x <- read.csv("activity.csv")
```

2.  Pre-processing data

```{r}
y <- x[is.na(x$steps)==FALSE,]
z <- tapply(y$steps,y$date,sum)
```

What is mean total number of steps taken per day?
-------------------------------------------------
1.  Histogram of the total number of steps taken each day (without NAs)

```{r}
hist(z)
```
2.  Calculate and report the mean and median total number of steps taken per day
Mean - total numnber of steps taken per day
```{r}
mean(z,na.rm=TRUE)
```
Median - total number of steps taken per day
```{r}
median(z,na.rm=TRUE)
```
What is the average daily activity pattern?
-------------------------------------------
1.  Time series plot of 5 minute interval
```{r}
zi <- tapply(y$steps,y$interval,mean)
plot(unique(y$interval),zi,type="l")
```
2.  Max number of steps for the average across 5 minute intervals.
```{r}

zid <- as.data.frame(zi)
zid <- cbind(zid,rownames(zid))
zid[order(zid$"zi",decreasing=TRUE),][1,2]
```

Inputing missing values
-----------------------

Pre-processing for missing values

1. Number of missing values
```{r}
cc <- complete.cases(x)
length(cc[which(cc==FALSE)])
```
2. Filling in NAs with mean value across days for the 5 minute interval

- Pre-processing - Creating a separate data frame of all missing values 
```{r}
xna <- x[which(is.na(x$steps)),]
yintmean <- tapply(y$steps,y$interval,mean)
yimdf <- as.data.frame(yintmean)
yimdf <- cbind(yimdf,rownames(yimdf))
colnames(yimdf) <- c("intmean","interval")
xna1 <- merge(xna,yimdf,by.x="interval",by.y="interval")
```

- Adding all mean values to the steps column

```{r}
xna$steps <- xna1$intmean
```

3.  New data set with filled in missing values

```{r}
ynew <- rbind(y,xna)
```

4.  Hist, mean and median of new data set

```{r}
hist(tapply(ynew$steps,ynew$date,sum))
```

Mean of the new data set per day

```{r}
mean(tapply(ynew$steps,ynew$date,sum))
```

Median of the new data set per day

```{r}
median(tapply(ynew$steps,ynew$date,sum))
```

Differences in activity patterns between weekdays and weekends
--------------------------------------------------------------

1.  Creating a new factor variable
```{r}
wdyn <- weekdays(strptime(ynew$date,format = '%Y-%m-%d'))
wdf <- wdyn %in% c("Saturday", "Sunday")
```

2. Panel plot for weekday and weekend plots

- Creating data of means per time interval for both weekdays and weekends

```{r}
ynew <- cbind(ynew, wdf)


ynew1 <- split(ynew, ynew$wdf)
wdays <- as.data.frame(ynew1[1])
wend <- as.data.frame(ynew1[2])
wdmean <- tapply(wdays$FALSE.steps,wdays$FALSE.interval,mean)
wemean <- tapply(wend$TRUE.steps,wend$TRUE.interval,mean)
wd <- as.data.frame(wdmean)
we <- as.data.frame(wemean)
wd <- cbind(wd,c("weekday"))
we <- cbind(we,c("weekend"))

colnames(we) <- c("Number of Steps", "daytype")
colnames(wd) <- c("Number of Steps", "daytype")

meanfin <- rbind(we,wd)

meanfin <- cbind(meanfin,rownames(meanfin))
colnames(meanfin)[3] = "Interval"
```

- Using lattice to plot
```{r}

library(lattice)
xyplot(meanfin$"Number of Steps" ~ meanfin$"Interval" | meanfin$"daytype", data=meanfin,layout=c(1,2), type="l", xlab="Interval",ylab="Number of Steps")


```

